Cover letter for the IJRR submission \\ ``An efficient acyclic contact planner for multiped robots''

### 2017-01-23 cover2.tex
finished cover letter

First of all we thank you for the time spent on considering our original manuscript. Indeed,
%This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 special issue.
As discussed with Prof. Hollerbach, we hope to have reviewer continuity for the revision.
The cover letter then plays more the role of a rebuttal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
Therefore, we kindly ask that this cover letter be forwarded to the reviewers of the new manuscript.

@@

\section{Recall of main paper contributions}
The paper is concerned with the problem of planning acyclic contact movements. 
More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot. 
Using the terminology of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
Our claim is that the solution of our planner leads to a straight-forward resolution of the last sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
In particular, our method proposes interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
This practical result corresponds to a significant improvement of the state of the art, for example regarding the capabilities of acyclic contact planners demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.

Regarding this contribution, the reviews mostly criticized the structure of the paper (too verbose) and the unjustified claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.

In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to easily solving $\mathcal{P}_3$. 
All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an contact simulator.
We also executed one of them on the real robot.
This extension, along with the correction of the other remarks by the reviewers, have been integrated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.

@@

\section{Validation of the method (addressing  $\mathcal{P}_3$) }
In his review, Reviewer 2 stated that our plans are likely to be not executable, even in simulation.

@@

We were really surprised by the remark, in particular because we had already validated some of our plans.
The stair climbing scenario has been demonstrated on the real HRP-2 robot, while the standing up had been demonstrated in simulation. 
While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
We believe that, as Reviewer 1 notices, this observation was due to the fact that we did not clearly define what we mean by a contact plan (i.e. a discrete sequence of key contact posture with the corresponding contact locations).
We thank the reviewers for these comments, since it pointed out that what we believed to be an evidence must be carefully explained and justified.

A rigorous definition of a contact plan is now detailed in section 5.1.

@@

\section{Success rates of the method}
At this point, we believe it is important to clarify our position regarding the concerns of both reviewers about the success rates of our paper.
In the paper, we are providing the success rates of our contact planner. 
Such an empirical study has yet never been provided for other contact planners.
Our method is not always successful, as very often with probabilistic searches.
More precisely, the hypothesis taken by \mP1 may lead to an unfeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an unfeasible \mP3.
We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
Even in case of failure, the planner is based on random sampling that approximate probabilistic completeness, and can be restarted to obtain another candidate trajectory.
In practice, it can always find a feasible solution fast enough for online applications by re-planning upon failure.

@@

\section{Originality and interest of the work}
We strongly disagree with Reviewer 2 regarding the originality of the work when he states that we simply ``\textit{experiment with minor heuristic adjustments}''.
Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.
These works, including the one of Hauser, are limited by the computational complexity of considering the full dimension of the problem.
In practice, the underlying complexity imply that they cannot lead to implementation that can be used in practical scenarios, as emphasized by the lack of such contact solvers during the DRC (as discussed by Atkinson).

@@

\section{A clearer paper}
As requested, we have deeply shortened the paper. More importantly, we have done an important work of defining in simpler terms
the different concepts that we use in the paper, and reduced their number.
We believe that the paper is much clearer in its current form.

\section{Justification of the simplifying assumptions}
Regarding several comments from both reviewers, stating that our paper does not justify the fact that contact feasibility implies equilibrium feasibility,
we now have added a section that discusses in depth the assumption (Section 7). We clearly describe which hypothesis (we call it the quasi-flat hypothesis, referring
to the contacts in the configurations, rather than the vague cluttered term) are required to validate our approach, and analyze
the success rates we obtained in each scenario to justify that the assumption works.

\section{Comparison with other methods}
As noted by Reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
since they were addressing the complete problem. We now give the complete computation times and compare them with previous works,
clearly demonstrating that our method outperforms them by orders of magnitude (Section 6.4).

\section{On the static equilibrium of each configuration}
Reviewer 1 stated that ``\textit{the planned contact path is only discrete with static equilibrium considered for each
configuration, which is not very realistic''}. While we agree with the observation, we first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande. Works that do not make this assumption such as Mordatch and al. only offer local convergence and cannot address these scenarios (e.g. on the ``Polaris car egress'' scenario reported in our paper). Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).

\section{Actual car model of the DARPA Challenge}
As asked by Reviewer 2, we now use the Polaris car model for demonstrating our egress scenario.

@@

\subsection{Reviewer 1}
\noindent
\begin{longtable}{|p{21em}|p{21em}|}
\hline
The paper addresses an important problem of planning robot motion involving contacts. It extends the idea of planning acyclic contacts by decoupling the issue into (1) planning a guide path for the root of the robot, and (2) planning a discrete sequence of acyclic and static equilibrium contact configurations along the path, with an interpolation of motion between two consecutive contact postures in the sequence. The introduced planner considers both collision avoidance and maintenance of desired contacts and aims at real-time or interactive performance. 
&Thanks for the comment.
For the decoupled approach to work, the planned guide path must guarantee that a sequence of stable contact configurations is associated with it to execute the path with desired contacts and yet without undesired collisions; such a guide path is called “equilibrium feasible”. The issue is how to find such a guide path efficiently. One key assumption of the paper is that for cluttered environments, “most of the times” a guide path that is “contact feasible” (referring to that at least one end-effector is in contact with the environment and there is no undesired collision between the robot and the environment) is also “equilibrium feasible”. It further assumes that if the reachable space of the robot along the guide path is in collision with the environment but the robot root configuration is not in collision, called the “reachability condition”, then the guide path is “contact feasible”. With those assumptions, the paper treated guide path planning as a geometric problem and used an RRT planner to solve the problem. 
& We agree with this summary.
However, the question is whether the above key assumption (of 
“contact feasible” implies “equilibrium feasible”) is reasonable. The paper did not provide justification for the assumption, and the assumption does not seem intuitive. The paper used terms such as “most of the times” and “cluttered contact planning” without clear and quantifiable definitions. With the above assumptions, planning becomes quite straightforward, and the planning algorithm used is rather conventional. 
We modified the paper to more rigorously discuss these hypotheses. 
We also believe that the main justification of their validity is through statistical tests, as reported in the result section.\newline
See Section 6 for details. 
For the next problem of planning a sequence of contact configurations given a “contact feasible” root path, the paper further decomposed the problem by planning the configuration of one limb at a time, assuming the root configuration and the other limb configurations were fixed. For each limb, N sampled configurations were considered (sampled off-line), and they were indexed by the corresponding end-effector position and stored in an octree beforehand. 
Because all that was done off-line, it is not clear how N was decided. 
The on-line planning step would search through the stored limb configurations (sorted according to some heuristic), one limb at a time (it seems), to find an almost-contact configuration and create a feasible limb contact configuration out of it. Different limbs were considered in some deterministic order: the limb that was contact-free the longest was considered first, and contact configurations were generated as long as the limb could maintain contact.  It is a bit confusing when interpolated contact configurations would be added between two adjacent contact configurations. Was it “in the hope” to keep a limb in contact or change limbs in contact? 
--- I THINK WE MUST DISCUSS THAT HERE ---
The manipulator Jacobian for each limb was assumed full rank, and no singular configurations were considered. 
--- I THINK WE MUST DISCUSS THAT HERE ---
The main issue with the proposed planner is that assumptions (commented earlier) and decisions were made in the planner conveniently in order to simplify the problem but without much justification, and there is no guarantee of a solution. The entire approach is heuristic and somewhat ad hoc. 
We agree that our approach is mostly heuristic.
Heuristics are somehow underestimated in robotics, where optimality principles are often preferred (this is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications).

The main hypotheses of the papers are: the division of contact planning in three problems; the reachability condition approximating the equilibrium condition; and the heuristics for building the contact sequence (\mP2) along the guide path (\mP1).
We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the other being statistically asserted.

We agree that the method also requires some other minor choices, like the ones discussed by the reviewers. 
We do not believe that these minor technical choices are important. 
If other choices are made, similar results would likely be obtained as well.
To keep the paper as short as possible, we did not spent time explaining these choices.
Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
The source code of the planner is also available (url in the paper) in case the exact implementation is needed.

Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.

@@

\\ \hline % ----------------------------------------------------------------------------------
Another major limitation is that the planned contact path is only discrete with static equilibrium considered for each configuration, which is not very realistic. The paper also pointed out this limitation in the Conclusions.
&
Our contribution is on solving \mP1-\mP2 efficiently. 
We believe that this is the core of the contact planning problem, as argued in introduction.
Indeed, solving \mP3 in all the cases studied in the paper was quite straight-forward.

Clearly, dividing the complete problem in 3 subproblems will ease the solver and possibly obstruct some solutions. The question is whether this choice is a sane trade-off or not. 
We insist that this division is quite classical in the state of the art.
In particular, the only planner able to cope with this division (by Mordatch et al) only leads to local convergence.

In the new version, we asserted the realism of the contact sequences by implementing a solution to \mP3, outputing full trajectories that were tested in a contact simulator and on the real HRP-2 robot.

\\ \hline % ----------------------------------------------------------------------------------
The important problem of transitioning motion from one contact configuration to the next was left out as beyond the scope of the paper. 
&
--- I THINK WE MUST DISCUSS THAT HERE ---
\\ \hline % ----------------------------------------------------------------------------------
The planner was tested in a number of diverse simulation examples, and there were related comparisons with existing work. This is commendable. However, there is no experimental result on a real robot. 
&
The contribution of the paper is on subproblems \mP1 and \mP2. 
We initially did not put robot results to emphasize this aspect and avoid confusing the reader.
Following the reviewer recommendations, we added a solver of subproblem \mP3, leading to complete movement, one of them being executed on the real robot. 
\newline See Section 2 for details.
\\ \hline % ----------------------------------------------------------------------------------
The writing of the paper can be much improved and made much more concise. The description of contact path planning in Section 4 is broken into different sub-sections, but not entirely in the order of how the sequence is generated, which is confusing. In Section 4.1, steps 3—6 talk about on-line generation, but the related decisions were introduced in Section 4.2.1 and 4.2.2, and none of those sub-sections has a title that clearly indicates the planning procedure. The paper has used unnecessary notations and symbols on the one hand, but vague languages such as “most of the time”, “cluttered problems”, “in the hope”, etc., on the other hand. Some symbols were not used consistently, for instance, m is used both to indicate mass and a direction. 
&
Thanks for carefully listing these problems. 
The paper was completely rewritten to make it shorter and more rigorous.
\\ \hline % ----------------------------------------------------------------------------------
\end{longtable}

@@

In the original paper, we claimed that any of these contact sequences would directly lead to a full contact trajectory (\mP3).
However, we did not asserted this claim. 
In the new version, we implemented a solver \mP3 and validated that all the contact sequences reported in the paper are leading to full contact trajectories, whose dynamic plausibility has been checked by execution on the real robot for one of them, and in a dynamic contact simulator for the others.

@@

In the revision, we have implement a solve \mP3 and converted all contact sequences in full contact trajectory, validated in simulation or on the real robot.
We must insist that this validation was straight-forward, on a scientific point of view. 
As initially stated, the produced contact sequences indeed lead directly to full trajectories.

@@

In terms of genericity, two key aspects must be considered.
First, our planner is good enough for 90\% of the situations that nowadays legged robot can meet.
Second, the general idea of reducing the contact planning problem is valid in any situations that we can imagine, and the heuristics and approximate models that we proposed can be adapted to extend the planner capabilities.
Automatic synthesis of these models is indeed a research direction that we are currently considering.

@@

It was indeed difficult to fully feel the enthusiasm of the reviewer, although we definitely appreciated his detailed comment and effort to improve the quality of the paper.
In conclusion, we truly believe that our solution answer to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.

@@

We hope that the revision of the paper will help the reviewer to meet our opinion.

>>>>

First of all we thank you for the time spent considering our original manuscript. Indeed,
%This cover letter for the manuscript ``An efficient acyclic contact planner for legged robots'' plays more the role of a rebutal, since the paper is a resubmission for the ISSR'15 sepcial issue, and that we ask for reviewer continuity.
this cover letter introduces the first revision of the manuscript ``An efficient acyclic contact planner for legged robots'', resubmitted for the ISSR'15 special issue.
As discussed with Prof. Hollerbach, we hope to have reviewer continuity for the revision.
The cover letter then plays more the role of a rebuttal. We left the original cover letter at the end of this document, as well as the original ISRR reviews.
Therefore, we kindly ask that this cover letter be forwarded to the reviewers of the new manuscript.

@@

Furthermore, we included the integrality of the previous reviews, decomposed in an array. Each row contains a paragraph of the review, as well as our answer to the
specific paragraph.

\section{Rebutal}
\subsection{Main contributions of the paper}
More precisely, we propose a very efficient method to address what we believe is the core of the difficulty of the contact planning problem: discovering the sequence of key contact postures that would lead to a safe and stable contact trajectory of the robot. 
The paper is concerned with the problem of planning acyclic contact movements. 
Using the terminology of the paper, our contact planner addressed the two sub-problems $\mathcal{P}_1$ (finding a guide path for the robot root) and $\mathcal{P}_2$ (finding a sequence of stable contact postures and corresponding contact locations).
Our claim is that the solution produced by our planner leads to a feasible sub-problem $\mathcal{P}_3$ (finding the complete trajectory connecting the contact postures).
We demonstrate that our method provides interactive performances, i.e. the planner needs less time to compute the solution than what is needed for the robot to execute it.
This practical result corresponds to a significant improvement of the state of the art, for example regarding the capabilities of acyclic contact planners demonstrated by the humanoid teams during the Darpa Robotics Challenge last year.

Regarding this contribution, the reviews essentially criticized the structure of the paper (too verbose) and the undemonstrated claim that the contact sequence output by the planner leads to a feasible resolution of $\mathcal{P}_3$.

In the revision, we implemented a solver inspired by the state of the art to solve $\mathcal{P}_3$ and empirically proved the claim that our $\mathcal{P}_1$-$\mathcal{P}_2$ planner leads to a feasible $\mathcal{P}_3$. 
All the robot contact sequences reported in the paper have been converted to full contact trajectories using this $\mathcal{P}_3$ solution, that were validated in an inverse dynamics simulator.
We also executed one of them on the real robot.
This extension, along with the correction of the other remarks by the reviewers, have been integrated in the revision paper, that we also reworked to make it clearer, shorter and sharper (the core of the paper is now 13 pages long, as opposed to 17 before) than the original manuscript.

@@

\subsection{Validation of the method (addressing  $\mathcal{P}_3$) }
In his review, Reviewer 2 stated that our plans are likely to be not executable, even in simulation.

@@

We were really surprised by the remark, in particular because we had already validated some of our plans.
The stair climbing scenario has been demonstrated on the real HRP-2 robot, while the standing up had been demonstrated in simulation. 
While it was mentioned in the paper, we did not insist on these demonstrations because our main contribution focuses on the other problems. 
We believe that, as Reviewer 2 notices, his observation is due to the fact that we did not clearly define what we mean by a contact plan (i.e. a discrete sequence of key contact posture with the corresponding contact locations).
We thank the reviewers for these comments, which pointed out that we had to carefully explain and justify this aspect of the paper.
A rigorous definition of a contact plan is now detailed in section 5.1.

@@

\subsection{Success rates of the method}
At this point, we believe it is important to clarify our position regarding the concerns of both reviewers about the success rates of our paper.
In the paper, we are providing the success rates of our contact planner. 
Such an empirical study has yet never been provided for other contact planners.
Our method is not always successful, as very often with probabilistic searches.
More precisely, the hypothesis taken by \mP1 may lead to an unfeasible \mP2 problem, and similarly \mP2 hypotheses may lead to an unfeasible \mP3.
We accept this fact and argue that it is mostly successful (i.e. more than 77\% of success in reasonable ``DRC-like'' scenarios).
Even in case of failure, since the planner is based on random sampling it can be restarted to obtain another candidate trajectory.
In practice, it can always find a feasible solution fast enough for online applications by re-planning upon failure.

@@

\subsection{Originality and interest of the work}
\label{orig}
We strongly disagree with Reviewer 1 and 2 regarding the originality of the work as ``\textit{heuristic adjustments}''.
Bretl formalized well the problem without providing an algorithm applicable to the general case. Escande focused on the contact planning $\mathcal{P}_2$. 
Our work tackles $\mathcal{P}_1$ as well, in a much more efficient manner than Bouyarmane et al.
These works, including the one of Hauser, are limited by the computational complexity of considering the full dimension of the problem.
In practice, the underlying complexity imply that they cannot lead to implementation that can be used in practical scenarios, as emphasized by the lack of such contact solvers during the DRC (as discussed by Atkinson).

@@

\subsection{A clearer paper}
As requested by both reviewers, we have deeply shortened the paper. More importantly, we have done an important work of defining in simpler terms
the different concepts that we use in the paper, and reduced their number.
We believe that the paper is much clearer in its current form.

More specifically, Reviewer 1 stated that the section related to the contact generation, in particular, was confusing. We rewrote Section 2.2 and Section 5 to give a clearer
description of the method, especially regarding the addition of intermediate contact configurations.

\subsection{Justification of the simplifying assumptions}
Regarding several comments from both reviewers, stating that our paper does not justify the fact that contact feasibility implies equilibrium feasibility,
we now have added a section that discusses in depth the assumption (Section 7). We clearly describe which hypothesis (we call it the quasi-flat hypothesis, referring
to the contacts in the configurations, rather than the vague cluttered term) are required to validate our approach, and analyze
the success rates we obtained in each scenario to justify that the assumption works.

\subsection{Comparison with other methods}
As noted by Reviewer 2, in our previous paper it was indeed complex to compare our work with previous contributions,
since they were addressing the complete problem. We now give the complete computation times and compare them with previous works,
clearly demonstrating that our method outperforms them by orders of magnitude (Section 6.4).

\subsection{Actual car model of the DARPA Challenge}
As asked by Reviewer 2, we now use the Polaris car model for demonstrating our egress scenario.

@@

\subsection{Reviewer 1}
\noindent
\begin{longtable}{|p{21em}|p{21em}|}
\hline
The paper addresses an important problem of planning robot motion involving contacts. It extends the idea of planning acyclic contacts by decoupling the issue into (1) planning a guide path for the root of the robot, and (2) planning a discrete sequence of acyclic and static equilibrium contact configurations along the path, with an interpolation of motion between two consecutive contact postures in the sequence. The introduced planner considers both collision avoidance and maintenance of desired contacts and aims at real-time or interactive performance. 
&Thanks for the comment.
\\ \hline % ----------------------------------------------------------------------------------
For the decoupled approach to work, the planned guide path must guarantee that a sequence of stable contact configurations is associated with it to execute the path with desired contacts and yet without undesired collisions; such a guide path is called “equilibrium feasible”. The issue is how to find such a guide path efficiently. One key assumption of the paper is that for cluttered environments, “most of the times” a guide path that is “contact feasible” (referring to that at least one end-effector is in contact with the environment and there is no undesired collision between the robot and the environment) is also “equilibrium feasible”. It further assumes that if the reachable space of the robot along the guide path is in collision with the environment but the robot root configuration is not in collision, called the “reachability condition”, then the guide path is “contact feasible”. With those assumptions, the paper treated guide path planning as a geometric problem and used an RRT planner to solve the problem. 
& We agree with this summary.
\\ \hline % ----------------------------------------------------------------------------------
However, the question is whether the above key assumption (of 
“contact feasible” implies “equilibrium feasible”) is reasonable. The paper did not provide justification for the assumption, and the assumption does not seem intuitive. The paper used terms such as “most of the times” and “cluttered contact planning” without clear and quantifiable definitions. With the above assumptions, planning becomes quite straightforward, and the planning algorithm used is rather conventional. 
&
We modified the paper to more rigorously discuss these hypotheses. 
We also believe that the main justification of their validity is shown by the statistical tests we provide in the result section.\newline
See Section 6 for details. 
\\ \hline % ----------------------------------------------------------------------------------
For the next problem of planning a sequence of contact configurations given a “contact feasible” root path, the paper further decomposed the problem by planning the configuration of one limb at a time, assuming the root configuration and the other limb configurations were fixed. For each limb, N sampled configurations were considered (sampled off-line), and they were indexed by the corresponding end-effector position and stored in an octree beforehand. 
Because all that was done off-line, it is not clear how N was decided. 
&
As explained in the paper, in our previous work (Tonneau et al. 2014), we discussed in details  how to choose N. We chose to refer to the paper rather than giving details  because of the space constraints of
the paper. In our experiments,
we always set $N=10000$.

\\ \hline % ----
The on-line planning step would search through the stored limb configurations (sorted according to some heuristic), one limb at a time (it seems), to find an almost-contact configuration and create a feasible limb contact configuration out of it. Different limbs were considered in some deterministic order: the limb that was contact-free the longest was considered first, and contact configurations were generated as long as the limb could maintain contact.  It is a bit confusing when interpolated contact configurations would be added between two adjacent contact configurations. Was it “in the hope” to keep a limb in contact or change limbs in contact? 
& 
The main reason to add intermediate contact configuration
is to respect the conditions imposed by our planner: at most one contact must be broken between two configurations. Sometimes the kinematic constraints are such
that for a given discretization step, several contacts have to be broken at once. Adding intermediate states allows to reposition one contact before the other.
Concretely, when an intermediate contact is added, one contact is broken and relocated. This is the role of \textsc{RepositionContacts}, as explained in Appendix D. We clarified this aspect in section 5.1.

\\ \hline % ----------------------------------------------------------------------------------
The manipulator Jacobian for each limb was assumed full rank, and no singular configurations were considered. 
&
For locomotion tasks, we are interested in avoiding the singular configurations that are not desirable.
Indeed, in our approach we discard these singular configurations, at the sampling phase, or by favoring non singular configurations
using our manipulability based heuristics.

However, this choice is not a limitation of our method: if singular configurations were required for some reason, the method would be able
to handle them.

\\ \hline % ----------------------------------------------------------------------------------
The main issue with the proposed planner is that assumptions (commented earlier) and decisions were made in the planner conveniently in order to simplify the problem but without much justification, and there is no guarantee of a solution. The entire approach is heuristic and somewhat ad hoc. 
&
As detailed in Section~\ref{orig} of this cover letter, we disagree regarding the fact that our approach is purely heuristic.
We claim that the dimensionality reduction proposed to address \mP1 is an important contribution to the field.

The main hypotheses of the papers are: the division of contact planning in three problems; the reachability condition approximating the equilibrium condition; and the heuristics for building the contact sequence (\mP2) along the guide path (\mP1).
We believe that these main hypotheses are properly discussed, some being theoretically justified (reachability condition), the others being statistically asserted.

%~ Then, regarding the heuristics we used to generate contacts and solve \mP2, we believe that heuristics are somehow underestimated in robotics, where optimality principles are often preferred. This is not the case in other fields, like discrete optimization, data routing or multi-agent diagnosis in web applications.
We agree that the method also requires some other minor choices, especially regarding \mP2, like the ones discussed by the reviewers. 
We do not believe that these minor technical choices are important. 
If other choices are made, similar results would likely be obtained as well.
To keep the paper as short as possible, we did not spent time explaining these choices.
%~ Any reader that would try to re-implement our approach would find enough details to reproduce our choices.
%~ The source code of the planner is also available (url in the paper) in case the exact implementation is needed.
%~ 
%~ Providing more details about any single technical choice would require 10 more pages, which is in contraction to the request of the reviewer to have the paper less verbose.

@@

\\ \hline % ----------------------------------------------------------------------------------
Another major limitation is that the planned contact path is only discrete with static equilibrium considered for each configuration, which is not very realistic. The paper also pointed out this limitation in the Conclusions.
&
We first recall that this assumption is made in the previous works that address cluttered scenarios, such as the work from Bretl and Escande. Works that do not make this assumption such as Mordatch and al. only offer local convergence and cannot address these scenarios (e.g. on the ``Polaris car egress'' scenario reported in our paper). Furthermore, this limitation is in fact less restrictive in our approach that the one of Escande, since while our plan is quasi static, our resulting motion is dynamic, as now explained in the paper (Section 5.1).

In the new version, we asserted the realism of the contact sequences by implementing a solution to \mP3, outputing full trajectories that were tested in a contact simulator and on the real HRP-2 robot.
See Sections 2 and 6 for details.

\\ \hline % ----------------------------------------------------------------------------------
The important problem of transitioning motion from one contact configuration to the next was left out as beyond the scope of the paper. 
The planner was tested in a number of diverse simulation examples, and there were related comparisons with existing work. This is commendable. However, there is no experimental result on a real robot. 
&
The contribution of the paper is on subproblems \mP1 and \mP2. 
We initially did not put robot results to emphasize this aspect and avoid confusing the reader.
Following the reviewer recommendations, we added a solver of subproblem \mP3, leading to complete movement, one of them being executed on the real robot. 
\newline See Section 2 for details.
\\ \hline % ----------------------------------------------------------------------------------
The writing of the paper can be much improved and made much more concise. The description of contact path planning in Section 4 is broken into different sub-sections, but not entirely in the order of how the sequence is generated, which is confusing. In Section 4.1, steps 3—6 talk about on-line generation, but the related decisions were introduced in Section 4.2.1 and 4.2.2, and none of those sub-sections has a title that clearly indicates the planning procedure. The paper has used unnecessary notations and symbols on the one hand, but vague languages such as “most of the time”, “cluttered problems”, “in the hope”, etc., on the other hand. Some symbols were not used consistently, for instance, m is used both to indicate mass and a direction. 
&
Thanks for carefully listing these problems. 
The paper was completely rewritten to make it shorter and more rigorous.
\\ \hline % ----------------------------------------------------------------------------------
\end{longtable}

@@

In the original paper, we claimed that our contact sequences would directly lead feasible problem (\mP3), and thus a feasible trajectory.
However, we did not assert this claim. 
In the new version, we implemented a solver \mP3 and validated that all the contact sequences reported in the paper are leading to full contact trajectories, whose dynamic plausibility has been checked by execution on the real robot for one of them, and in a dynamic contact simulator for the others.

@@

In the revision, we have implement a solver for \mP3 and converted all contact sequences to whole-body trajectories, validated in simulation or on the real robot.
%~ We must insist that this validation was straight-forward, on a scientific point of view. 
As initially stated, the produced contact sequences indeed lead directly to complete trajectories.

@@

In terms of genericity, two key aspects must be considered.
First, our planner is good enough for the majority of situations that nowadays legged robot can meet.
Second, the general idea of reducing the contact planning problem is valid in any situations that we can imagine, and the heuristics and approximate models that we proposed can be adapted to extend the planner capabilities.
Automatic synthesis of these models is indeed a research direction that we are currently considering.

@@

%~ It was indeed difficult to fully feel the enthusiasm of the reviewer, although
We definitely appreciated the detailed comments and effort to improve the quality of the paper.
In conclusion, we truly believe that our solution answers to the main difficulty of contact planning, by building a realistic contact sequence which easily leads to the construction of a complete contact trajectory, and behind, the execution of it on the real robot.

@@

We hope that the revision of the paper will help the reviewers to meet our opinion.
